---
title: "**R Code Notes**"
author: "*created by Sylvia Kim*"
date: "*version 2022-02-05*"
output:
 html_document:
    toc: true
    toc_float:
      toc_collapsed: true
    toc_depth: 3
    number_sections: false
    theme: yeti
---
***

## **Introduction**
For the purpose of self study and code review, this R markdown report has been created. The original contents can be found on [**Sylvia's GitHub.**](https://github.com/selgik/R-practice) 

### **Package used**
* dplyr  
* janitor  
* palmerpenguins  
* skimr  
* tidyverse  

### **Table of contents**
#### **Part 1: Concepts**  
1. Functions and variables  
2. Vectors and data frames  

#### **Part 2: Basic**
1. Install and load packages 
2. Prepare data 
3. Inspect data
4. Organize data
5. Manipulate data
6. Nested and pipe
7. Logical operator and conditional statement
8. Visualize data
9. Save the visuals

#### **Part 3: Tips**
1. Do not assume
2. Check for bias

*****

## **Part 1: Concepts**  

### **1-1. Functions and variables**  
**functions**: a code to perform a specific task. Type function and then add argument within the (). Example:
```{r}
print("It's a beautiful day!")
```


**variable**: a value which can be stored for later use. Also called as objects. Define variable name followed by <- and function.
```{r}
var_x <- "this is variable"
var_y <- 123.45
```

A calculation can be performed with variable as below:
```{r}
var_y - 23.45
```

Ask about function by adding ? before function name.
```{r eval=FALSE}
?print()
```

or find out more about the packages.
```{r eval=FALSE}
browseVignettes("tidyverse")
```


### **1-2. Vectors and data frames**  
**vector**: a group of data elements of the same type stored in a sequence. Once a vector is created, it will show as a set of data.

```{r}
vec_1 <- c(12,34,56,78.9)
vec_2 <- c(1L, 5L, 10L)
vec_3 <- c("Anna", "Beta", "Cera", "Delta")
```

now, let's run one of vector (vec_3) below.
```{r echo=FALSE}
vec_3
```

**list**: it's similar to vector but it can contain different data type. 
```{r}
list_1 <- list("a", 1L, 3.5, FALSE)
```

Other operations can be done to vectors including:  
calculating the length of vector 
```{r}
length(vec_3)
```

or, assigning titles to vector can be done too.
```{r echo=TRUE}
names(vec_1) <- c("spring", "summer", "fall", "winter")
print(vec_1)
```

The same can be done to list.
```{r}
list1 <- list('x-axis'=1, 'y-axis'=2, 'z-axis'=3)
print(list1)
```


**data frame**: collection of columns, typically imported from different source.
Example below:
```{r}
df_1 <- data.frame(city=c("NY", "SF", "CO"), days=c(2.4, 4.4, 5.1), rank=c(2,1,3))
```


or codes can be added step by step by first defining the variables:
```{r eval=FALSE}
city <- c("NY", "SF", "CO")
days <- c(2.4, 4.4, 5.1)
rank <- c(2,1,3)
df_1 <- data.frame(city,days,rank)
```

Now, let's run the code to see the result.
```{r echo=FALSE}
df_1
```

**matrix**: two-dimentional collection of data elements, containing a single data type.

```{r}
matrix(c(3:10), nrow=2)
matrix(c(3:10), ncol=2)
```

Note that *matrix(c(3:9), nrow=2)* will give an error, as 7 elements are not 2x multiplier.

***  
## **Part 2: Basic**
### **2-1. Install and load packages**
Check installed packages. If needed, install for use.
```{r eval=FALSE}
installed.packages()
install.packages("palmerpenguins")
```

load package and/or dataset before using it. Nothing will show up in the console:
```{r eval=FALSE}
library(palmerpenguins) 
data(penguins)
```

### **2-2. Prepare data**
Import csv file and select certain columns:
```{r eval=FALSE}
csv_file <- read_csv("test.csv")
select(csv_file, column1, column2, column3)
```

Import excel file and read specific sheet:
```{r eval=FALSE}
read_excel("test.xls")
excel_sheets("test.xls")
read_excel("test.xls", sheet="sales")
```


### **2-3. Inspect data**
View will display data in reader-friendly table format in separate tab:
```{r eval=FALSE}
View(penguins)  
```

Also, summary data can be reviewed with below functions. 
```{r include=FALSE}
library("palmerpenguins")
library("tidyverse")
```

```{r eval=FALSE}
head(penguins)
colnames(penguins)
glimpse(penguins)
str(penguins)
```

As an example, str() has been run as below:
```{r}
str(penguins)
```

Summary statistics can also be checked:
```{r eval=FALSE}
skim_without_charts(penguins) 
summary(penguins)                  
```

Example of summary():
```{r}
summary(penguins)                  
```

Inspect specific column(s):
```{r eval=FALSE}
select(penguins, species) 
```

By adding - sign before the column, result will display all columns except for marked column. See example, same as *select(penguins, -species)*: 
```{r}
penguins %>%
  select(-species)
```


### **2-4. Organize data**

Sort the data. By adding - sign, sort in DESC order instead of default ASC:
```{r eval=FALSE}
arrange(penguins, bill_length_mm)
arrange(penguins, desc(bill_length_mm))
penguins %>% arrange(bill_length_mm)
penguins %>% arrange(-bill_length_mm)
```

Filter values:
```{r eval=FALSE}
filter(penguins, species=='Gentoo')
```

Example of using arrange and filter functions:
```{r}
penguins %>% 
  filter(species=="Gentoo") %>%
  select(species, island, body_mass_g) %>%
  arrange(-body_mass_g)
```

Find max, min, mean: if dataset contains NA, result will show as NA.
```{r eval=FALSE}
min(penguins$year)
max(penguins$year)
mean(penguins$year)
```

Group data for summary statistics. Below example illustrates the process of grouping data -> removing NA values -> assigning new column name -> summarizing them.
```{r warning=FALSE}
penguins %>% 
  group_by(species, island) %>% drop_na() %>%
  summarize(mean_bl=mean(bill_length_mm),
            max_bl=max(bill_length_mm))
```

Save cleaned data frame:
```{r eval=FALSE}
cleaned_penguins <- penguins %>% arrange(bill_length_mm)
cleaned2_penguins <- penguins %>% select(island, species)
```


### **2-5. Manipulate data**
Rename column or variable: rename(dataset, new_name=old_name): 
```{r eval=FALSE}
rename(penguins, weight=body_mass_g)
```

Update all columns to upper/lower case
```{r eval=FALSE}
rename_with(penguins,toupper)
rename_with(penguins,tolower)
```

Clean column names by ensuring only characters, numbers and _ are in the columns
```{r eval=FALSE}
clean_names(penguins)
```

Combine columns:
```{r eval=FALSE}
unite(data_set, 
      'new_column_name', 
      coulmn1_to_unite, 
      column2_to_unite, 
      sep=' ,')
```

Applying as below:
```{r eval=FALSE}
example <- bookings_data %>%
  select(arrival_date_year, arrival_date_month) %>%
  unite(arrival_year_month, c("arrival_date_year",
                              "arrival_date_month"), sep = " ,")
```

Separate column:
```{r eval=FALSE}
separate(data_set, 
         column_to_separate, 
         into=c('column1', 'colmn2'), 
         sep= ' ')
```

Add column: mutate(dataset, new_column=explain)
```{r eval=FALSE}
mutate(penguins, body_mass_kg=body_mass_g/1000)
```

Example below:
```{r}
penguins %>%
  mutate(body_mass_kg=body_mass_g/1000) %>%
  select(species, island, body_mass_kg) %>%
  arrange(-body_mass_kg)
```

Summary statistics: summarize(dataset, col_name=mean(col))
```{r eval=FALSE}
penguins %>%
  drop_na() %>%
  summarize(avg_g=mean(body_mass_g), sum_g=sum(body_mass_g))
```

### **2-6. Nested query and pipe**
Nested query can be written as below:
```{r eval=FALSE}
arrange(filter(ToothGrowth, dose==0.5), len)
```

Using pipe, process step is clearer and less cluttered:
```{r eval=FALSE}
filtered_toothgrowth <- ToothGrowth %>%
  filter(dose==0.5) %>%
  arrange(len)
```


### **2-7. Conditional statement**
Logical operators: and &, or |, not!
```{r}
x <-10

x<12 & x>11      # false as 10 is not bigger than 11.
x<12 | x>11      # true, less than 12 or bigger than 11.
!x>11            # true, not bigger than 11.
!(x>15 | x<5)    # true, not bigger than 15 or less than 5
```

Conditional statement: if(){then}
```{r}
x <-5

if(x>0){
  print("x is a positive number")
  } else {
    print("x is a negative number")
    }
```


```{r}
y <-1982

if (y>1990) {
  print("Group1")
  } else if (y>1980) {
    print("Group2")
    } else {
      print("Group3")
      }
```

### **2-8. Visualize data**  
ggplot(data)+geom_shape(mapping=aes(argument)). geom_bar will display bar chart, geom_point will display scatter plot. See below:
```{r warning=FALSE}
ggplot(data=penguins)+ 
  geom_point(mapping=aes(x=flipper_length_mm, y=body_mass_g))
```

Above can be also written as below:
```{r eval=FALSE}
ggplot(data=penguins, 
       mapping=aes(x=flipper_length_mm,y=body_mass_g))+
  geom_point()
```

Enhance the chart by mapping color, size, shape or alpha(difference in density):
```{r warning=FALSE}
ggplot(data=penguins)+
  geom_point(mapping=aes(x=flipper_length_mm,
                        y=body_mass_g,
                        color=species,
                        shape=species))
```

See the chart with alpha aes:
```{r warning=FALSE}
ggplot(data=penguins)+
  geom_point(mapping=aes(x=flipper_length_mm,
                         y=body_mass_g,
                         alpha=species))
```


Below is the alternative way of writing codes for archive purpose:
```{r eval=FALSE}
ggplot(data=penguins, aes(x=flipper_length_mm,y=body_mass_g,color=species))+
  geom_point()

ggplot(data=penguins, aes(x=flipper_length_mm,y=body_mass_g,color=species))+
  geom_point(color="purple")

ggplot(data=penguins, aes(x=flipper_length_mm,y=body_mass_g,color=species))+
  geom_point()+
  facet_wrap(~species)
```

If color needs to be applied to entire chart instead of specific variables: 
```{r warning=FALSE}
ggplot(data=penguins)+
  geom_point(mapping=aes(x=flipper_length_mm,y=body_mass_g),
             color="purple")
```

Different chart types can be created by changing geom. For example, smooth line as below:
```{r warning=FALSE}
ggplot(data=penguins)+
  geom_smooth(mapping=aes(x=flipper_length_mm,y=body_mass_g))
```

bar chart below. If color is used instead of fill, the outline will only be colored:
```{r warning=FALSE}
ggplot(data=penguins)+
  geom_bar(mapping=aes(x=species,fill=species)) 
```

stacked bar chart:
```{r warning=FALSE}
ggplot(data=penguins)+
  geom_bar(mapping=aes(x=species,fill=island)) 
```

Examine relationship between trend line and data points by adding two geom:
```{r warning=FALSE}
ggplot(data=penguins)+
  geom_smooth(mapping=aes(x=flipper_length_mm, y=body_mass_g))+
  geom_point(mapping=aes(x=flipper_length_mm, y=body_mass_g))
```

There are also different types of soothing lines. For reference:
```{r eval=FALSE}
ggplot(data=penguins,aes(x=flipper_length_mm,y=body_mass_g))+
  geom_smooth(method="loess")                                       #Loess smoothing
  
ggplot(data=penguins,aes(x=flipper_length_mm,y=body_mass_g))+
  geom_smooth(method="gam",formula=y~s(x))                          #Gam smoothing
```

Facet function is also handy for focusing on specific data points:
```{r warning=FALSE}
ggplot(data=penguins)+
  geom_point(mapping=aes(x=flipper_length_mm, y=body_mass_g))+
  facet_wrap(~species)
```

facet_grid for 1+ facet:
```{r warning=FALSE}
ggplot(data=penguins)+
  geom_point(mapping=aes(x=flipper_length_mm,y=body_mass_g,color=species))+
  facet_grid(sex~species)
```

Other functions can be used as normal:
```{r eval=FALSE}
penguins %>%
  filter(island=="Biscoe") %>%
  ggplot(mapping=aes(x=flipper_length_mm,y=body_mass_g,color=species))+
  geom_point()
```

Adding label and annotation will help readers understand about data:
```{r warning=FALSE}
ggplot(data=penguins)+
  geom_point(mapping=aes(x=flipper_length_mm,y=body_mass_g, color=species))+
  labs(title="Palmer Penguins: Relationship between Body Mass vs Flipper Length",
       subtitle="Body Mass (gram) / Flipper Length (mm) ",
       caption="R Package used: palmerpenguins")+
  annotate("text", x=220,y=4000,label="positive relationship observed",
           angle=45, fontface="bold", size=3)
```

Label can be added using variables too:
```{r eval=FALSE}
minamount <- min(dataframe$column)
maxamount <- max(dataframe$column)
  
labs(caption=paste0("The minimum is ",minamount," and the maximum is ", maxamount))
```


Add x/y axis header title if missing:
```{r eval=FALSE}
labs(x="X-axis name", y="Y-axis name")
```

Rotate x-axis headers:
```{r eval=FALSE}
theme(axis.text.x=element_text(angle=45))
```

### **2-9. Save the visuals**

ggsave will save the latest visual:
```{r eval=FALSE}
ggsave("test_visuals.png", width=5, height=5)
```

***  

## **Part 3: Tips** 
### **3-1. Do not assume**

Summary may indicate the similar dataset for below groups:
```{r}
library('Tmisc')
data(quartet)
quartet %>%
  group_by(set) %>%
  summarize(mean(x),sd(x),mean(y),sd(y),cor(x,y))
```


But visualizing them will reveal the difference between the grouped data:
```{r warning=FALSE}
ggplot(quartet,aes(x,y))+
  geom_point()+
  geom_smooth(method=lm,se=FALSE)+
  facet_wrap(~set)
```

### **3-2. Check for bias**
Package needed: 
```{r include=FALSE}
library(SimDesign)
library(dplyr)
```

```{r eval=FALSE}
install.packages("SimDesign")
library("SimDesign")
```

Compare data:
```{r}
actual_data <- c(10,20,30,40,50)
predicted_data <-c(8,14,22,39,45)
bias(actual_data,predicted_data)
```

Another example:
```{r}
actual_data2 <- c(10,20,30,40,50)
predicted_data2 <- c(12,24,39,47,55)
bias(actual_data2, predicted_data2)
```

The more the result is closer to 0, the less the data is biased.


### *END*
*If you have any feedback, please do not hesitate to reach out sylviahk416@gmail.com*
